Generating Empirically Optimized Composed Matrix Kernels from MATLAB Prototypes

### 2008-12-18 doc/papers/iccs09/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@387 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
Gail's edits to abstract
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@401 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

% ICCS'09 http://www.iccs-meeting.org/iccs2009/
% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/iccs09
% Important dates:
% Full papers submission (extended)       December 20, 2008
% Notification of acceptance of papers    February 2, 2009
% Camera ready papers                        February 15, 2009
% Max 10 pages, llncs style

@@

% first the title is needed
\title{Generating Empirically Optimized Composed Matrix Kernles from MATLAB Prototypes}

@@

\author{Elizabeth Jessup\inst{1} 
  \and Boyana Norris\inst{2}
  \and Albert Hartono\inst{3}
  \and Jeremy Siek\inst{1}}

\authorrunning{Jessup, Norris, Hartono, Siek}

\institute{University of Colorado at Boulder \\
  \email{\{elisabeth.jessup,jeremy.siek\}@colorado.edu}
  \and Argonne National Laboratory \\
  \email{\{norris\}@mcs.anl.gov}
  \and Ohio State University \\
  \email{\{hartonoa\}@cse.ohio-state.edu}}

\tocauthor{Elizabeth Jessup, Boyana Norris, Albert Hartono, and Jeremy Siek}
\maketitle

@@

%\vspace{-.2in}
\subsection{A MATLAB Compiler}
\label{sec:matlab}
%\input{matlab}
In this section we describe the design and operation of the MATLAB to C compiler we have developed~\cite{Siek}.    Figure~\ref{fig:compiler} gives an overview of the compilation process. 
%
\begin{wrapfigure}{r}{2in}
\vspace{-.2in}
\centering
\includegraphics[width=2in]{figures/compile.png}
\caption{The MATLAB compiler.}
\label{fig:compiler}
\vspace{-.3in}
\end{wrapfigure}
The MATLAB kernel specification is parsed into a high-level intermediate
representation in the form of a dataflow graph.  This dataflow graph is then
iteratively processed until all of the implementation choices have been made.
The compilation process consists of three phases: analysis, refinement, and
optimization, which are together iterated until all of the implementation
decisions have been made.  The graph is then translated into C code.
%\subsubsection{The Dataflow Graph.}
%
An example dataflow graph for the GEMVER kernel, defined below, is shown
in Figure~\ref{fig:gemver-dataflow}.
Each node represents a parameter of the kernel or an operation.  The arrows
indicate the flow of data. At first the graph specifies what operations are
to be performed but does not contain any implementation details. The symbol
$\times$ in the depicted graph stands variously for outer product,
matrix-vector multiplication, and scalar-vector multiplication, and it does
not yet specify, for example, whether the outer products compute a row or a
column at a time of the result matrix.
\begin{figure}[ht]
\begin{minipage}{.3\textwidth}
\begin{align*}
  A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
  x &\gets \beta A^T y + z \\[-0.5ex]
  w &\gets \alpha A x
\end{align*} 
\end{minipage}
\begin{minipage}{.6\textwidth}
\includegraphics[width=3in]{figures/gemver-dataflow.png}
\end{minipage}
\caption{Dataflow graph (right) for the GEMVER kernel (left).}
\label{fig:gemver-dataflow}
\end{figure}

@@

\begin{figure}[tb]
\vspace{-.1in}
\begin{center}
\includegraphics[width=.65\textwidth]{figures/orio.png}
\end{center}
\vspace{-.25in}
\caption{Overview of the Orio empirical tuning process.}
\label{fig:orio}
\end{figure}

@@

\begin{table}[htb] 
\vspace{-.1in} 
\caption{Composed BLAS operations used in our experiments}
\label{tbl:blas-ops}
\small
\centering 
\vspace{-.1in} 
\begin{tabular}{|c|c|c|c|} 
\hline 
Name & Input & Output & C code from MATLAB compiler \\
\hline 
\hline 
VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ 
\hline 
ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ 
\hline 
 & ${\bf A}$,$a$,$b$, & ${\bf B}$, & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
& $\overrightarrow{y}$,$\overrightarrow{z}$ & & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\
\hline 
BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
& & & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\
\hline 
\end{tabular} 
\vspace{-.2in} 
\end{table}

@@

\begin{minipage}[b]{.5\textwidth}
\includegraphics[height=2in,width=\textwidth]{figures/vadd_bgp.png}
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\includegraphics[height=2in,width=\textwidth]{figures/atax.png}
\end{minipage}\\

@@

\begin{minipage}[b]{.5\textwidth}
\includegraphics[height=2in,width=\textwidth]{figures/gemver.png}
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\includegraphics[height=2in,width=\textwidth]{figures/bicgkernel.png}
\end{minipage}\\

@@

The experiments of the remaining operations were performed on the
multicore Intel Xeon. Included in these experiments are performance
numbers for six code variants: the C code generated by the MATLAB
compiler (''C from MATLAB''), three BLAS-based implementations that
use Intel MKL, ATLAS, and the default BLAS library on Ubuntu 8.04, and
the sequential and parallel code variants tuned by Orio (``Orio
(Seq.)'' and ``Orio (Par.)'', respectively).

@@

>>>>

% ICCS'09 http://www.iccs-meeting.org/iccs2009/}
% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/iccs09
% Important dates:
% Full papers submission (extended)       December 20, 2008
% Notification of acceptance of papers    February 2, 2009
% Camera ready papers                        February 15, 2009
% Max 10 pages, llncs style

@@

% first the title is needed
\title{Generating Empirically Optimized Composed Matrix Kernels from MATLAB Prototypes}

@@

\author{Boyana Norris\thanks{The work of this author was supported by the Office of Advanced Scientific Computing Research, Office of Science, U.S. Dept. of Energy, under Contract DE-AC02-06CH11357.}\inst{1}
  \and Albert Hartono\inst{2}
  \and Elizabeth Jessup\thanks{The work of this author was supported by the
  National Science Foundation under grants no. CCF-0430646 and CCF-830458.}\inst{3}
  \and Jeremy Siek\inst{3}
%
\authorrunning{Norris, Hartono, Jessup, Siek}
%
\institute{Argonne National Laboratory \\
  \email{norris@mcs.anl.gov}
  \and Ohio State University \\
  \email{hartonoa@cse.ohio-state.edu}
  \and University of Colorado at Boulder \\
  \email{\{elisabeth.jessup,jeremy.siek\}@colorado.edu}}}

\tocauthor{B. Norris, A. Hartono, E. Jessup, and J. Siek}
\maketitle

@@

%\vspace{-.2in}
\subsection{A MATLAB Compiler}
\label{sec:matlab}
%\input{matlab}
In this section we describe the design and operation of the MATLAB to C compiler we have developed~\cite{Siek}.    Figure~\ref{fig:compiler} gives an overview of the compilation process. 
%
\begin{wrapfigure}{r}{2in}
\vspace{-.2in}
\centering
\includegraphics[width=2in]{figures/compile.png}
\caption{The MATLAB compiler.}
\label{fig:compiler}
\vspace{-.2in}
\end{wrapfigure}
The MATLAB kernel specification is parsed into a high-level intermediate
representation in the form of a dataflow graph.  This dataflow graph is then
iteratively processed until all of the implementation choices have been made.
The compilation process consists of three phases: analysis, refinement, and
optimization, which are together iterated until all of the implementation
decisions have been made.  The graph is then translated into C code.
%\subsubsection{The Dataflow Graph.}
%
An example dataflow graph for the GEMVER kernel, defined below, is shown
in Figure~\ref{fig:gemver-dataflow}.
Each node represents a parameter of the kernel or an operation.  The arrows
indicate the flow of data. At first the graph specifies what operations are
to be performed but does not contain any implementation details. The symbol
$\times$ in the depicted graph stands variously for outer product,
matrix-vector multiplication, and scalar-vector multiplication, and it does
not yet specify, for example, whether the outer products compute a row or a
column at a time of the result matrix.
\begin{figure}[ht]
\begin{minipage}{.3\textwidth}
\begin{align*}
  A &\gets A + u_1 v_1^T + u_2 v_2^T \\[-0.5ex]
  x &\gets \beta A^T y + z \\[-0.5ex]
  w &\gets \alpha A x
\end{align*} 
\end{minipage}
\begin{minipage}{.6\textwidth}
\includegraphics[width=3in]{figures/gemver-dataflow.png}
\end{minipage}
\caption{Dataflow graph (right) for the GEMVER kernel (left).}
\label{fig:gemver-dataflow}
\vspace{-.2in}
\end{figure}

@@

\begin{figure}[tb]
\vspace{-.3in}
\begin{center}
\includegraphics[width=.65\textwidth]{figures/orio.png}
\end{center}
\vspace{-.25in}
\caption{Overview of the Orio empirical tuning process.}
\label{fig:orio}
\vspace{-.2in}
\end{figure}

@@

\begin{table}[htb] 
\vspace{-.1in} 
\caption{Composed BLAS operations used in our experiments.}
\label{tbl:blas-ops}
\small
\centering 
%\vspace{-.1in} 
\begin{tabular}{|c|c|c|c|} 
\hline 
Name & Input & Output & C code from MATLAB compiler \\ 
\hline 
\hline 
%& & & \\
VADD & $\overrightarrow{w}$,$\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$ &
$\overrightarrow{x} = \overrightarrow{w} + \overrightarrow{y} + \overrightarrow{z}$ \\ [1ex]
\hline 
%& & & \\
ATAX & $A$,$\overrightarrow{x}$ & $\overrightarrow{y}$ & 
$\overrightarrow{y} = A' * (A * \overrightarrow{x})$ \\ [1ex]
\hline 
%& & & \\
 & ${\bf A}$,$a$,$b$, &  & ${\bf B} = {\bf A} + \overrightarrow{u_1} * \overrightarrow{v_1}' + \overrightarrow{u_2} * \overrightarrow{v_2}'$\\
GEMVER & $\overrightarrow{u_1}$,$\overrightarrow{u_2}$,$\overrightarrow{v_1}$,$\overrightarrow{v_2}$, & ${\bf B}$,  & $\overrightarrow{x} = b * ({\bf B}' * \overrightarrow{y}) + \overrightarrow{z}$\\
& $\overrightarrow{y}$,$\overrightarrow{z}$ & $\overrightarrow{x}$,$\overrightarrow{w}$ & $\overrightarrow{w} = a * ({\bf B} * \overrightarrow{x})$\\ [1ex]
\hline 
& & & $\overrightarrow{q} = {\bf A} * \overrightarrow{p}$ \\
BiCG Kernel & ${\bf A}$,$\overrightarrow{p}$,$\overrightarrow{r}$ & $\overrightarrow{q}$,$\overrightarrow{s}$ & $\overrightarrow{s} = {\bf A}' * \overrightarrow{r}$\\ [1ex]
\hline 
\end{tabular} 
%\vspace{-.2in} 
\end{table}

@@

\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/vadd_bgp.png}
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/atax.png}
\end{minipage}\\

@@

\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.png}
\end{minipage}
&
\begin{minipage}[b]{.5\textwidth}
\includegraphics[width=\textwidth]{figures/bicgkernel.png}
\end{minipage}\\

@@

The experiments of the remaining operations were performed on the
multicore Intel Xeon. Included in these experiments are performance
numbers for six code variants: the C code generated by the MATLAB
compiler (``C from MATLAB''), three BLAS-based implementations that
use Intel MKL, ATLAS, and the default BLAS library on Ubuntu 8.04, and
the sequential and parallel code variants tuned by Orio (``Orio
(Seq.)'' and ``Orio (Par.)'', respectively).

@@

%\newpage
%\vfill
%\begin{flushright}
%\scriptsize
%\framebox{\parbox{2.4in}{The submitted manuscript has been created by
%UChicago Argonne, LLC, Operator of Argonne National Laboratory
%("Argonne").  Argonne, a U.S. Department of Energy Office
%of Science laboratory, is operated under Contract No.
%DE-AC02-06CH11357.  The U.S. Government retains for itself, and
%others acting on its behalf, a paid-up, nonexclusive, irrevocable
%worldwide license in said article to reproduce, prepare derivative works,
%distribute copies to the public, and perform publicly and display
%publicly, by or on behalf of the Government.}}
%\normalsize
%\end{flushright}
